On finiteness of structures produced deterministically by co-transcriptional folding

### 2018-11-22 ttt.tex
update 11/22 in tokyo

If a bead is stabilized through a tunnel section, then it can provide some bonds. Let us consider bonds of $C_i$. $C_i$ is represented $C = (W,P,H)$ where $|W| = i + n$ $C_i$ contains $\alpha \cdot (i + n) - 2|H|$ bonds because $C_i$ consists of $i + n$ beads and a bead has just $\alpha$ bonds and then $2 |H|$ of the those bonds are already used. However, even if a bead has an available bond, $w[j]$ $(j > i)$ might not be able to use this bond because the bond has possibility that it is blocked by transcripts $w[i+1..j-1]$. Number of $binding\ capabilities$ does not contain that case so that it is at most $\alpha \cdot (i + n) - 2|H|$.

\begin{definition}[binding capability]
We say that $p$ is $reachable$ if there exists directed path $P^\prime$ from $q$ to $p$ such that the path of $P^\prime$ does not cross the path of $P$ where $q_0$ is a bead which is latest stabilized, and $q_m$ is a reachable point. Let $B_i$ be $( \{ (h, i) | {}^\forall h < i \} \cup \{ (i, j) | {}^\forall j > i \}) \cap H$. Moreover, let $R_i$ be a set of inside $\hexagon_{w_i}^1$ that are free and reachable.
 We represent the number of binding capabilities of a conformation $C_j$ as $\#bc(C_j)$. $\#bc(C_j)$ is defined by $\sum^{j}_{k=-n+1} \min \{ |B_k|, |R_k|  \}$.
\end{definition}

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \fill (0:0) circle [radius=0.05];
      \fill (0:1.2) circle [radius=0.05];
      \fill (180:1.2) circle [radius=0.05];
      \node[above] at (0:1.2) { $p^\prime$ };
      \node[above] at (0:0) { $p$ };
      \node[above] at (180:1.2) { $p^\prime$ };

      \fill (60 : 1.2) circle [radius=0.1];
      \fill (-60 : 1.2) circle [radius=0.1];
      \fill (120 : 1.2) circle [radius=0.1];
      \fill (-120 : 1.2) circle [radius=0.1];
      \node[above] at (60 :1.2) { $a_{j_1}$ };
      \node[above] at (-60 :1.2) { $a_{j_2}$ };
      \node[above] at (120 :1.2) { $a_{j_3}$ };
      \node[above] at (-120 :1.2) { $a_{j_4}$ };

      \begin{scope}[shift=(0:5)]
        \fill (0:0) circle [radius=0.1];
        \fill (0:1.2) circle [radius=0.05];
        \fill (180:1.2) circle [radius=0.1];
        \node[above] at (180:1.2) { $a_{i-2}$ };
        \node[above] at (0:0) { $a_{i-1}$ };
        \draw[->] (180:1.1) -- (180:0.1);

        \fill (60 : 1.2) circle [radius=0.1];
        \fill (-60 : 1.2) circle [radius=0.1];
        \fill (120 : 1.2) circle [radius=0.1];
        \fill (-120 : 1.2) circle [radius=0.1];
        \node[above] at (60 :1.2) { $a_{j_1}$ };
        \node[above] at (-60 :1.2) { $a_{j_2}$ };
        \node[above] at (120 :1.2) { $a_{j_3}$ };
        \node[above] at (-120 :1.2) { $a_{j_4}$ };
      \end{scope}
      \begin{scope}[shift=(0:10)]
        \fill (0:0) circle [radius=0.1];
        \fill (0:1.2) circle [radius=0.1];
        \fill (180:1.2) circle [radius=0.1];
        \node[above] at (180:1.2) { $a_{i-2}$ };
        \node[above] at (0:0) { $a_{i-1}$ };
        \node[above] at (0:1.2) { $a_i$ };
        \draw[->] (180:1.1) -- (180:0.1);
        \draw[->] (0:0.1) -- (0:1.1);

        \fill (60 : 1.2) circle [radius=0.1];
        \fill (-60 : 1.2) circle [radius=0.1];
        \fill (120 : 1.2) circle [radius=0.1];
        \fill (-120 : 1.2) circle [radius=0.1];
        \node[above] at (60 :1.2) { $a_{j_1}$ };
        \node[above] at (-60 :1.2) { $a_{j_2}$ };
        \node[above] at (120 :1.2) { $a_{j_3}$ };
        \node[above] at (-120 :1.2) { $a_{j_4}$ };
      \end{scope}
    \end{tikzpicture} 
    \caption{Through a tunnel section}
    \label{TTT_tunnel_intro}
  \end{center}
\end{figure}

\begin{theorem}[Tunnel Troll Theorem]
Let $\Xi$ be a unary oritatami system of $\delta = 1, \alpha \geq 2$. If there is a part of transcript $w[i..j+1] = bttt...tb$, then $\#bc(C_i)$ is more than $\#bc(C_j)$.
\end{theorem}

@@

>>>>

%If a bead is stabilized through a tunnel section, then it can provide some bonds. Let us consider bonds of $C_i$. $C_i$ is represented $C = (W,P,H)$ where $|W| = i + n$ $C_i$ contains $\alpha \cdot (i + n) - 2|H|$ bonds because $C_i$ consists of $i + n$ beads and a bead has just $\alpha$ bonds and then $2 |H|$ of the those bonds are already used. However, even if a bead has an available bond, $w[j]$ $(j > i)$ might not be able to use this bond because the bond has possibility that it is blocked by transcripts $w[i+1..j-1]$. Number of $binding\ capabilities$ does not contain that case so that it is at most $\alpha \cdot (i + n) - 2|H|$.

We say that point $p$ is reachable from a conformation $C$ if there exists a directed path $P^\prime$ from the last point of $C$ that does not cross the path of $C$. We define $binding\ capability$ with reachable.

\begin{definition}[binding capability]
Let $B_i$ be $( \{ (h, i) | {}^\forall h < i \} \cup \{ (i, j) | {}^\forall j > i \}) \cap H$. Moreover, let $R_i$ be a set of neighbors of $w[i]$ that are free and reachable from $C_j$ where $C_j$ is a conformation which stabilized until $w[j]$.
 We represent the number of binding capabilities of a conformation $C_j$ as $\#bc(C_j)$. $\#bc(C_j)$ is defined by $\sum^{j}_{k=-n+1} \min \{ |B_k|, |R_k|  \}$.
\end{definition}

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \draw (0:0) circle [radius=0.05];
      \draw (0:1) circle [radius=0.05];
      \draw (180:1) circle [radius=0.05];
      \node[above] at (0:1) { $p^\prime$ };
      \node[above] at (0:0) { $p$ };
      \node[above] at (180:1) { $p^\prime$ };

      \fill (60 : 1) circle [radius=0.1];
      \fill (-60 : 1) circle [radius=0.1];
      \fill (120 : 1) circle [radius=0.1];
      \fill (-120 : 1) circle [radius=0.1];
      \node[above] at (60 :1) { $a_{j_1}$ };
      \node[above] at (-60 :1) { $a_{j_2}$ };
      \node[above] at (120 :1) { $a_{j_3}$ };
      \node[above] at (-120 :1) { $a_{j_4}$ };

      \begin{scope}[shift=(0:5)]
        \fill[blue] (0:0) circle [radius=0.1];
        \draw (0:1) circle [radius=0.05];
        \fill (180:1) circle [radius=0.1];
        \node[above] at (180:1.2) { $a_{i-2}$ };
        \node[above] at (0:0) { $a_{i-1}$ };
        \draw[->, blue] (180:0.9) -- (180:0.1);

        \fill (60 : 1) circle [radius=0.1];
        \fill (-60 : 1) circle [radius=0.1];
        \fill (120 : 1) circle [radius=0.1];
        \fill (-120 : 1) circle [radius=0.1];
        \node[above] at (60 :1) { $a_{j_1}$ };
        \node[above] at (-60 :1) { $a_{j_2}$ };
        \node[above] at (120 :1) { $a_{j_3}$ };
        \node[above] at (-120 :1) { $a_{j_4}$ };
      \end{scope}
      \begin{scope}[shift=(0:10)]
        \fill (0:0) circle [radius=0.1];
        \fill[blue] (0:1) circle [radius=0.1];
        \fill (180:1) circle [radius=0.1];
        \node[above] at (180:1) { $a_{i-2}$ };
        \node[above] at (0:0) { $a_{i-1}$ };
        \node[above] at (0:1) { $a_i$ };
        \draw[->] (180:0.9) -- (180:0.1);
        \draw[->,blue] (0:0.1) -- (0:0.9);

        \fill (60 : 1) circle [radius=0.1];
        \fill (-60 : 1) circle [radius=0.1];
        \fill (120 : 1) circle [radius=0.1];
        \fill (-120 : 1) circle [radius=0.1];
        \node[above] at (60 :1) { $a_{j_1}$ };
        \node[above] at (-60 :1) { $a_{j_2}$ };
        \node[above] at (120 :1) { $a_{j_3}$ };
        \node[above] at (-120 :1) { $a_{j_4}$ };
      \end{scope}
    \end{tikzpicture} 
    \caption{Through a tunnel section}
    \label{TTT_tunnel_intro}
  \end{center}
\end{figure}

\begin{theorem}[Tunnel Troll Theorem]
Let $\Xi$ be a unary oritatami system of $\delta = 1, \alpha \geq 2$. If there are indices $i$ and $j$ such that $S[i..j+1] = bt^{j-i}b$, then $\#bc(C_i)$ > $\#bc(C_j)$.
\end{theorem}

@@

\subsubsection{$\alpha = 3$}

\begin{theorem}[$\delta = 1, \alpha = 3$]
Let $\Xi$ be a unary oritatami system of $\delta = 1, \alpha = 3$. It can yield only a finite structure whose size is $\mathcal{O}(n)$.
\end{theorem}

\begin{lemma}
\label{TTT_a3_2b_lemma}
 Let $p$ be a point whose neighbors is occupied at least two point. If $w[i]$ is not stabilized and $w[i-1]$ includes neighbors of $p$, then $w[i]$ is stabilized at $p$ with at least one bond, $w[i]$ is stabilized at another point of $p$ otherwise with at least two bond except any neighbors of $p$ is occupied.
\end{lemma}

\begin{proof}[proof of lemma]
Assume the transcript is stabilized until $w[i-1]$. One of neighbors of $p$ is not $w[i-1]$ where this bead regards $n_1$. If $w[i-1]$ include neighbors of $p$ and $w[i]$ is stabilized at another point of $p$ with one bond. Then, any neighbors do not have bond without $w[i-1]$. Neighbors of $n_1$ have to be occupied at least five according to lemma \ref{TTT_neighbor_lemma} and two of them include neighbors of $p$ where each of them regards $n_2, n_3$. In the same way, five neighbors of $n_2$ and $n_3$ are occupied and each of one of them includes neighbors of $p$ where they regard $n_4, n_5$. one of $n_5$'s neighbors includes neighbors of $p$ where it regards $n_6$. Then, any neighbors of $p$ are occupied. That is if some neighbors of $p$ are free, then there exists a bead which has bonds in neighbors.
\end{proof}

\begin{proof}
Let us show that $\#bc(C_{i-1}) > \#bc(C_i)$, that is when $w[i]$ is stabilized, $w[i]$ uses at least two hands.
Let us assume $w[i]$ is able to be stabilized with using one hand. Fig.\ref{TTT_a3_w} exhibits all the three kinds of possibility of stabilized $w[i]$. Then, $w[i]$ can be also stabilized at $n_3$.

\begin{paragraph}{Case of straight}
\begin{itemize}
\item[-] Case of $n_3$ is free\\
According to assumption, $w[i]$ uses only one hand. Therefore, any neighbors of $n_3$ are occupied according to the lemma\ref{TTT_a3_2b_lemma}. $n_3$ and the point which is stabilized $w[i]$ are free so that $n_1$ has some bond by lemma\ref{TTT_neighbor_lemma}. Accordingly, this situation is non-deterministic. Thus, $n_3$ and $n_4$ have to be occupied because of symmetry.

\item[-] Otherwise\\
Because of  $S[i] = b$, at least one of $n_1$ and $n_2$ have to be free. Let us regard that $n_1$ is free. Neighbors of $n_1$ have to be occupied and at least two neighbors of $n_{-1}$ have to be free for $n_1$ and $w[i]$. According to lemma\ref{TTT_neighbor_lemma}, $n_{-1}$ have some hand. Therefore $w[i]$ can be also stabilized $n_1$ that is this situation is non-deterministic. Thus, one of $n_3$ and $n_4$ has to be free.
\end{itemize}

Therefore, this case is false.
\end{paragraph}

\begin{paragraph}{Case of obtuse}
\begin{itemize}
\item[-] Case of $n_3$ is free\\
Any neighbors of $n_3$ have to be occupied but the point which is stabilized $w[i]$ is free. Thus $n_3$ has to be occupied.
\item[-] Case of $n_4$ is free\\
%n4??????????lemma2??n2???????lemma???n0??????????????w[i]???????????????????????bead?????????????1?????
%??n0?????????????n3???????????lemma1??n-3????????n-3???????????n-2??????????n-1???????????????w[i]???????????????????w[i]????????????
%??n3?????????????n0????????????lemma1??n-1???????????n-2?n-3???????w[i]?????????????
%???n0 n3???beead???????????n0 n3 ????????????lemma1??n-1 n-3????????n-1,n-3????????????n-3???????????
%???n-4????????????in order to provide inifinite structures?
\item[-] Case of $n_2$ is free\\
Any neighbors of $n_2$ have to be occupied so that $n_0$ is occupied. Any neighbors of $n_0$ except $n_2$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_2$ has to be occupied.
\item[-] Case of $n_0$ is free\\
Any neighbors of $n_0$ have to be occupied so that $n_{-1}$ is occupied. Any neighbors of $n_{-1}$ except $n_0$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_0$ has to be occupied. 
\end{itemize}

Therefore, any situations contradict $S[i] = b$.

\end{paragraph}

\begin{paragraph}{Case of acute}
\begin{itemize}
\item[-] Case of $n_4$ is free\\
$n_4$ and a point which is stabilized $w[i]$ are free so that  $w[i-2]$ has some hands according to lemma\ref{TTT_neighbor_lemma}.
However, $w[i]$ can be also stabilized $n_4$ in this case. Thus, $n_4$ has to be occupied.
\item[-] Case of $n_2$ is free\\
According to lemma\ref{TTT_a3_2b_lemma}, $n_0$ has to be occupied. $n_1$ has to be also occupied because of lemma\ref{TTT_neighbor_lemma}.
We consider this case just like case of obtuse and that $n_4$ is free. Then if $w[i-2]$ binds $w[i]$, any $n_{-1}, n_{-2}, n_{-3}$ are occupied. If $n_1$ binds $w[i]$, this case is same. Also if $n_1$ and $w[i-2]$ do not have any hand, any $n_{-1}, n_{-2}, n_{-3}$ are occupied. Therefore, $w[i+1]$ cannot be provided.
\item[-] Case of $n_0$ is free\\
Any neighbors of $n_0$ have to be occupied so that $n_{1}$ is occupied. Any neighbors of $n_{1}$ except $n_0$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_0$ has to be occupied. 
\item[-] Case of $n_1$ is free\\
Any neighbors of $n_1$ have to be occupied so that $n_{-1}$ is occupied. Any neighbors of $n_{-1}$ except $n_1$ have to be also occupied but the point which is stabilized $w[i]$ is free. Thus $n_1$ has to be occupied. 
\end{itemize}

Therefore, any situations contradict $S[i] = b$.
\end{paragraph}\\

Hence, assumption that w[i] is able to be stabilized with using one hand is false.
Therefore, when w[i] is stabilized, w[i] uses at least two hands.
\end{proof}

\begin{figure}
  \begin{center}
  \begin{tabular}{c c c}
 \begin{minipage}{0.3\hsize}
 \centering
    \begin{tikzpicture}

      \fill[shift=(180:1)] (0,0) circle [radius=0.1];
      \fill[shift=(180:0)] (0,0) circle [radius=0.1];

      \fill[blue](0:1) circle [radius=0.05];

      \draw[->] (180:0.9) -- (180:0.1);
      \draw[->, blue] (0:0.1) -- (0:0.9);

	\node[above] at (180:1) {$w_{[i-2]}$};
	\node[below] at (180:0) {$w_{[i-1]}$};
	\node[above] at (0:1) {$w_{[i]}$};

	\foreach \theta in {60,-60,120,-120}{
   	   \draw [shift=(\theta:1)] (-0.05,-0.05) rectangle (0.05,0.05);
  	}
	 \draw [shift=(60:1)] (0.95,-0.05) rectangle (1.05,0.05);
 	\node[above] at (60:1) {$n_1$};
	\node[below] at (-60:1) {$n_2$};
	\node[above] at (120:1) {$n_3$};
	\node[below] at (-120:1) {$n_4$};
	\node[above, shift=(0:1)] at (60:1) {$n_{-1}$};
    \end{tikzpicture}
    \end{minipage}

      \fill[shift=(180:1)] (0,0) circle [radius=0.1];
      \fill[shift=(180:0)] (0,0) circle [radius=0.1];

      \fill[blue](60:1) circle [radius=0.05];

      \draw[->] (180:0.9) -- (180:0.1);
      \draw[->, blue] (60:0.1) -- (60:0.9);

	\node[above] at (180:1) {$w_{[i-2]}$};
	\node[below] at (180:0) {$w_{[i-1]}$};
	\node[above] at (60:1) {$w_{[i]}$};

	\foreach \theta in {0,-60,120,-120}{
   	   \draw [shift=(\theta:1)] (-0.05,-0.05) rectangle (0.05,0.05);
  	}
	\draw [shift=(60:1)] (0.95,-0.05) rectangle (1.05,0.05);
	\draw [shift=(60:1), shift=(60:1)] (-0.05,-0.05) rectangle (0.05,0.05);
	\draw [shift=(60:1), shift=(120:1)] (-0.05,-0.05) rectangle (0.05,0.05);

 	\node[right] at (0:1) {$n_0$};
	\node[below] at (-60:1) {$n_2$};
	\node[above] at (120:1) {$n_3$};
	\node[below] at (-120:1) {$n_4$};
	\node[above, shift=(0:1)] at (60:1) {$n_{-1}$};
	\node[above, shift=(60:1)] at (60:1) {$n_{-2}$};
	\node[above, shift=(60:1)] at (120:1) {$n_{-3}$};
    \end{tikzpicture}
    \end{minipage}

      \fill[shift=(180:1)] (0,0) circle [radius=0.1];
      \fill[shift=(180:0)] (0,0) circle [radius=0.1];

      \fill[blue](120:1) circle [radius=0.05];

      \draw[->] (180:0.9) -- (180:0.1);
      \draw[->, blue] (120:0.1) -- (120:0.9);

	\node[above] at (180:1) {$w_{[i-2]}$};
	\node[below] at (180:0) {$w_{[i-1]}$};
	\node[above] at (120:1) {$w_{[i]}$};

	\foreach \theta in {0,60,-60,-120}{
   	   \draw [shift=(\theta:1)] (-0.05,-0.05) rectangle (0.05,0.05);
  	}
	\draw [shift=(120:1), shift=(60:1)] (-0.05,-0.05) rectangle (0.05,0.05);
	\draw [shift=(120:1), shift=(120:1)] (-0.05,-0.05) rectangle (0.05,0.05);
	\draw [shift=(120:1), shift=(180:1)] (-0.05,-0.05) rectangle (0.05,0.05);

 	\node[right] at (0:1) {$n_0$};
	\node[above] at (60:1) {$n_1$};
	\node[below] at (-60:1) {$n_2$};
	\node[below] at (-120:1) {$n_4$};
	\node[above, shift=(120:1)] at (60:1) {$n_{-1}$};
	\node[above, shift=(120:1)] at (120:1) {$n_{-2}$};
	\node[above, shift=(120:1)] at (180:1) {$n_{-3}$};
    \end{tikzpicture}
    \end{minipage}
    \end{tabular}
    \caption{All possible directions of $w[i]$: straight, obtuse, acute.}
    \label{TTT_a3_w}
  \end{center}
\end{figure}

\subsubsection{$\alpha = 4$}
\begin{theorem}[$\delta = 1, \alpha = 4$]
Let $\Xi$ be a unary oritatami system of $\delta = 1, \alpha = 4$. It can yield only a finite structure whose size is $\mathcal{O}(n^2)$.
\end{theorem}

\begin{lemma}
\label{TTT_a4_2b_lemma}
Any beads which are already stabilized by some bonds use at least two bonds.
\end{lemma}

\begin{proof}[proof of lemma]
Let us consider when $w[i]$ is stabilized by only one bond. See Fig.\ref{TTT_a4_lemma_fig}. According to lemma\ref{TTT_neighbor_lemma}, if $n_3$ is free, $w[i-2]$ has some hands. Thus, $n_4$ has to be occupied in order to stabilize deterministically. Moreover, also $n_2$ has to be occupied for deterministic and also $n_0, n_1$. $n_1$ has some hands because $n_3$ is free. Therefore, $w[i]$ is stabilized at $n_3$ and it has to use at least two hands. It contradict assumption.
\end{proof}

\begin{proof}
According to lemma\ref{TTT_a4_2b_lemma}, when $w[i]$ is stabilized, it has to use at least two bonds. Let us consider when a bead $w[i]$ which is the first bead out of $\hexagon_{w[-n+1]}^n$ is stabilized. See Fig.\ref{TTT_a4_first}. any $n_0, n_1, n_3, n_5$ is free because if some of them is occupied, $w[i]$ is not the first bead out of $\hexagon_{w[-n+1]}^n$. At least two neighbors of $w[i]$ except predecessor have to be occupied in order to bind. In this case, a point which is able to put a bead is only $n_2$. Therefore, any transcript cannot be stabilized in out of $\hexagon_{w[-n+1]}^n$. Hence oritatami system can yield only a finite structure whose size is $\mathcal{O}(n^2)$ in $\delta =1, \alpha = 4$.
%lemma???a=4??#bc(C_i) >= #bc(C_{i+1})? ?upper?left
\end{proof}

\begin{figure}
 \centering
    \begin{tikzpicture}
	\fill[shift=(180:1)] (0,0) circle [radius=0.1];
      \fill[shift=(180:0)] (0,0) circle [radius=0.1];

      \draw[->] (180:0.9) -- (180:0.1);

	\node[above] at (180:1) {$w_{[i-2]}$};
	\node[below] at (180:0) {$w_{[i-1]}$};
	\node[above] at (0:1) {$n_0$};

	\foreach \theta in {0,60,-60,120,-120}{
   	   \draw [shift=(\theta:1)] (-0.05,-0.05) rectangle (0.05,0.05);
  	}
 	\node[above] at (60:1) {$n_1$};
	\node[below] at (-60:1) {$n_2$};
	\node[above] at (120:1) {$n_3$};
	\node[below] at (-120:1) {$n_4$};
    \end{tikzpicture}
    \caption{$\alpha = 4$: when $w[i]$ is stabilized}
    \label{TTT_a4_lemma_fig}
\end{figure}

\begin{figure}
 \centering
    \begin{tikzpicture}
	\fill[shift=(-120:1)] (0,0) circle [radius=0.1];
      \fill[shift=(180:0), blue] (0,0) circle [radius=0.05];

      \draw[dashed] (180:2) -- (0:2);
      \draw[->, blue] (-120:0.9) -- (-120:0.1);

	\node[below] at (-120:1) {$w_{[i-1]}$};
	\node[above] at (180:0) {$w_{[i]}$};
	\node[above] at (0:1) {$n_0$};

	\foreach \theta in {0,60,-60,120,180}{
   	   \draw [shift=(\theta:1)] (-0.05,-0.05) rectangle (0.05,0.05);
  	}

 	\node[above] at (60:1) {$n_1$};
	\node[below] at (-60:1) {$n_2$};
	\node[above] at (120:1) {$n_3$};
	\node[above] at (180:1) {$n_5$};

	\node at (0:2.5) {$n+1$};
    \end{tikzpicture}
    \caption{the first bead out of $\hexagon_{w[-n+1]}^n$}
    \label{TTT_a4_first}
\end{figure}
